Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2019-11-21 manuscript/climate_sensitivity_draft.Rmd
Updated with author feedback to end of 1st paragraph of discussion
#65

- We analyzed tree-ring records for twelve species of an oak-hickory forest - representing 97% of woody productivity - in the 25.6-ha ForestGEO plot in Virginia (USA) to determine how tree size, microhabitat, and species' traits interactively shape drought responses across the three strongest regional droughts over a 60-year period (1950 - 2009).

@@

\newpage
### Introduction
**See comment 4,31**
Forests globally play a critical role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change [@friedlingstein_climatecarbon_2006]. An important aspect of this uncertainty lies with physiological responses of trees to drought [@kennedy_implementing_2019]. In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014; @cook_global_2014; @cook_unprecedented_2015; @mankin_mid-latitude_2019], sometimes despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015]. Droughts resulting from global change have been affecting forests worldwide [@allen_global_2010], and are expected to continue severely impacting forests [@anderegg_consequences_2013; @adams_temperature_2009; @breda_temperate_2006; @hartmann_research_2018; @allen_underestimation_2015; @abrams_interdisciplinary_2016; @adams_multi-species_2017]. During drought, larger trees generally suffer greater reductions in growth and increased mortality rates [e.g., @bennett_larger_2015; @stovall_tree_2019], resulting in disproportionate impacts on forest C storage [@meakem_role_2018]. As a result, forest drought responses stand to strongly impact forest feedbacks to climate change [@mcdowell_multi-scale_2016; @charney_observed_2016]. Yet, accurate characterization of drought responses remains a modeling challenge [@kennedy_implementing_2019]-- in part because some of the underlying mechanisms remain unclear (**REF**). Understanding forest responses to drought requires increased functional understanding of how tree size, microhabitat, and species' traits jointly confer individual-level vulnerability or resistance, and the extent to which their influence is consistent across droughts.

@@

To yield functional understanding of how tree size, microenvironment, and species' traits collectively shape drought responses, we test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and forest census data from a 25.6-ha ForestGEO plot in Virginia (USA). First, we focus on the role of tree size and its interaction with microenvironment. We confirm that, consistent with most forests globally, larger-diameter trees tend to have lower drought resistance in this forest, which is in an ecoregion **see comment 10** represented by only one study in [@bennett_larger_2015] (*H1.0*). 
We then test hypotheses designed to disentangle the relative importance of tree height (*H1.1*), crown exposure (*H1.2*), and soil water availability, which should be greater for larger trees in dry but not in perpetually wet microsites (*H1.3*). **see comments 11,12**. 
Second, we focus on the role of species' functional and hydraulic traits and their interaction with tree height. We hypothesize that drought resistance will be negatively affected by wood density (following observed patterns by **e.g.** @hoffmann_hydraulic_2011), and positively affected by specific leaf area and xylem architecture (*i.e.*, ring or diffuse/ semi-ring porous), but that hydraulic leaf traits including $PLA_{dry}$ and $\pi_{tlp}$ will prove better predictors (*H2.1*). 
We then test whether these traits correlate with tree height (*H2.2*), potentially driving the observed tendency for taller trees to suffer more during drought after accounting for trait effects (*H2.3*). 
Finally, we focused on variability among droughts, asking how community resistance varied across droughts (*H3.1*) and whether the factors confirming vulnerability or resistance varied across droughts (*H3.2*).

@@

Within or just outside the ForestGEO plot, we collected data on a suite of variables including tree size, microenvironment, and species traits (Table 2). Destructive sampling was only carried out within the plot. **see comment 14** The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols, whereby all free-standing woody stems $\ge$ 1cm diameter at breast height (DBH) were mapped, tagged, measured at DBH, and identified to species [@condit_tropical_1998]. From this census data, we used measurements of DBH from 2008 to calculate historical DBH and data for all stems $\ge$ 10cm to analyze functional trait composition relative to tree height (all analyses described below). Census data, which were last updated in 2019, are available through the ForestGEO data portal (www.forestgeo.si.edu).

@@

Our findings support the hypothesis *H2.3* that the observed tendency for larger trees to have greater growth reductions during drought (lower $R$) is driven by height itself, as opposed to more drought-sensitive traits in larger trees (Tables 1,5). As discussed above, there was little meaningful variation in traits with height at the community level. When $ln[H]$ and hydraulic traits were considered together in multivariate models, the effect of $ln[H]$ on $R$ was consistently negative (Table 5). Simultaneously in the 1999 univariate model, this co-consideration resulted in a non-signficant positive $ln[H]$-$R$ correlation (Table 4).

@@

We reject the hypothesis (*H3.1*) that overall community responses varied across droughts. Within the context of mixed effects models, there were no significant differences in $R$ across drought years (Table 4). This is consistent with the observation that the distribution of $R$ values was similar across droughts (Fig. 1b).

We mostly reject the hyopthesis (*H3.2*) that directions of responses varied across droughts. In the majority of cases, response directions were consistent across droughts in both univariate and multivariate models (Tables 1,4,5). However, there were a few exceptions--most commonly in the categorical variables ($CP$ and $XP$) but also for $PLA_{dry}$ in one multivariate model for the 1999 drought (Tables 4, 5). These differences may very well be random, as opposed to statistically meaningful. Among the univariate models, there was no instance where predictor variables signficantly improved the models of two different droughts (dAIC $\ge$ 2), but with contrasting coefficients (Table 4). Among the multivariate models, $CP$ was not consistently in the top models for any drought (Table 5), and $PLA_{dry}$ only appeared with a positive coefficient in the weakest of six models for the 1999 drought (contrasting with a negative coefficient in the univariate model; Table 4). The difference most likely to be real is that ring-porous species had lower $R$ than diffuse- and semi-ring- porous species in the 1977 drought, contrasting with higher resistance in 1966 and 1999 (Tables 4,5), but note that $XP$ was not a signficant predictor on its own for the 1977 drought.

We support the hypothesis (*H3.3*) that the strength of predictor variables was different across the droughts (Tables 1,4,5). For instance, $ln[H]$ and $PLA_{dry}$ had much stronger negative effects in 1966 than in the other two years, $ln[TWI]$ had the strongest negative effect in 1977, and $XP$ (higher $R$ among ring-porous trees) was strongest in 1999 (Tables 4,5).

@@

Our results reveal how tree size, microhabitat, and hydraulic traits shaped tree growth responses across three droughts in a temperate deciduous forest (Table 1). The tendencey for larger trees to suffer more, observed here as in forests around the world [@bennett_larger_2015], was driven primarily by their height. There was a marginal additional effect of crown exposure, with the most exposed and the most suppressed trees suffering most--consistent with observations of both greater drought sensitivity of exposed trees (e.g., [@suarez_factors_2004]; [@scharnweber_confessions_2019]) and greater sensitivity of suppressed and crowded individuals (**REFS**). There was no evidence that soil water availability increased drougth resistance; in contrast, trees in wetter topographic positions suffered more (consistent with [@zuleta_drought-induced_2017]), and larger rooting volume provided no advantage in the drier microenvironments. The lower drought resistance of larger trees was not driven by any tendency for the canopy to be dominated by more drought-sensitive species. Drought sensitivity was not consistently linked to species' $LMA$, $WD$, or xylem architecture. However, drought resistance was negatively correlated with the leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts, which is physiologically logical ([@scoffoni_leaf_2014];[@bartlett_correlations_2016]; [@medeiros_extensive_2019]) but scientifically novel in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses. The direction of responses was mostly consistent across droughts, indicating that they were driven by fundamental physiological mechanisms; however, the strengths of each predictor varied across droughts, indicating that specific drought characteristics interact with tree size, microenvironment, and traits to shape which individuals suffer most. These findings significantly advance our knowledge of the factors that confer vulnerability or resistance on trees during drought.

The droughts considered here were of similar severity (Fig. 1b) and fairly moderate; droughts of this magnitude have occurred with an average frequency of approximately one per 10-15 years (Fig. 1a, @helcoski_growing_2019). Therefore, we excpect that most species are adapted, and individual trees acclimatized, to survive droughts of this nature. While the majority of trees experienced reduced growth, a substantial portion had increased growth (Fig. 1b), underlining the fact that these droughts did not induce extreme stress on the entire forest. It is likely for this reason, combined with the fact that many factors other than climate affect tree growth in closed-canopy forests, that our best models characterize only a modest amount of variation: 11-13% for all droughts combined, and 21-26% for each individual drought (Table 5). Methodologically, the moderate nature of these droughts is an advantage because our analysis considers only trees that survived all of these droughts, and we lack information on the trees that were killed. These are likely to be relatively modest in number, and local forest monitoring data stretching back to the late 1980s confirms that the 1999 drought did not trigger major declines in tree abundance or biomass (*Anderson-Teixeira et al., in revision*). Thus, the droughts considered here are substantially weaker than those that have triggered massive tree die-off (e.g., [@allen_global_2010]), many of which have shaped our understanding about the role of tree size [@bennett_larger_2015; @stovall_tree_2019] and--to some extent--traits underlying drought response mechanisms (@greenwood_tree_2017; @anderegg_meta-analysis_2016). Nevertheless, our results are consistent with findings from more extreme droughts.

>>>>

- We analyzed tree-ring records for twelve species of an oak-hickory forest - representing 97% of woody productivity - in the 25.6-ha ForestGEO plot in Virginia (USA) to determine how tree size, microenvironment characteristics, and species' traits interactively shape drought responses across the three strongest regional droughts over a 60-year period (1950 - 2009).

@@

\newpage
### Introduction
**See comment 4,31**
Forests globally play a critical role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change [@friedlingstein_climatecarbon_2006]. An important aspect of this uncertainty lies with physiological responses of trees to drought [@kennedy_implementing_2019]. In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014; @cook_global_2014; @cook_unprecedented_2015; @mankin_mid-latitude_2019], sometimes despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015]. Droughts resulting from global change have been affecting forests worldwide [@allen_global_2010], and are expected to continue severely impacting forests [@anderegg_consequences_2013; @adams_temperature_2009; @breda_temperate_2006; @hartmann_research_2018; @allen_underestimation_2015; @abrams_interdisciplinary_2016; @adams_multi-species_2017]. During drought, larger trees generally suffer greater reductions in growth and increased mortality rates [e.g., @bennett_larger_2015; @stovall_tree_2019], resulting in disproportionate impacts on forest C storage [@meakem_role_2018]. As a result, forest drought responses stand to strongly impact forest feedbacks to climate change [@mcdowell_multi-scale_2016; @charney_observed_2016]. Yet, accurate characterization of drought responses remains a modeling challenge [@kennedy_implementing_2019]-- in part because some of the underlying mechanisms remain unclear (**REF**). Understanding forest responses to drought requires increased functional understanding of how tree size, microenvironment, and species' traits jointly confer individual-level vulnerability or resistance, and the extent to which their influence is consistent across droughts.

@@

To yield functional understanding of how tree size, microenvironment characteristics, and species' traits collectively shape drought responses, we test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and forest census data from a 25.6-ha ForestGEO plot in Virginia (USA). First, we focus on the role of tree size and its interaction with microenvironment. We confirm that, consistent with most forests globally, larger-diameter trees tend to have lower drought resistance in this forest, which is in an ecoregion **see comment 10** represented by only one study in [@bennett_larger_2015] (*H1.0*). 
We then test hypotheses designed to disentangle the relative importance of tree height (*H1.1*), crown exposure (*H1.2*), and soil water availability, which should be greater for larger trees in dry but not in perpetually wet microsites (*H1.3*). **see comments 11,12**. 
Second, we focus on the role of species' functional and hydraulic traits and their interaction with tree height. We hypothesize that drought resistance will be negatively affected by wood density (following observed patterns by **e.g.** @hoffmann_hydraulic_2011), and positively affected by specific leaf area and xylem architecture (*i.e.*, ring or diffuse/ semi-ring porous), but that hydraulic leaf traits including $PLA_{dry}$ and $\pi_{tlp}$ will prove better predictors (*H2.1*). 
We then test whether these traits correlate with tree height (*H2.2*), potentially driving the observed tendency for taller trees to suffer more during drought after accounting for trait effects (*H2.3*). 
Finally, we focused on variability among droughts, asking how community resistance varied across droughts (*H3.1*) and whether the factors confirming vulnerability or resistance varied across droughts (*H3.2*).

@@

Within or just outside the ForestGEO plot, we collected data on a suite of variables including tree size, microenvironment characteristics, and species traits (Table 2). Destructive sampling was only carried out within the plot. **see comment 14** The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols, whereby all free-standing woody stems $\ge$ 1cm diameter at breast height (DBH) were mapped, tagged, measured at DBH, and identified to species [@condit_tropical_1998]. From this census data, we used measurements of DBH from 2008 to calculate historical DBH and data for all stems $\ge$ 10cm to analyze functional trait composition relative to tree height (all analyses described below). Census data, which were last updated in 2019, are available through the ForestGEO data portal (www.forestgeo.si.edu).

@@

Our findings support Hypothesis *H2.3* that the observed tendency for larger trees to have greater growth reductions during drought (lower $R$) is driven by height itself, as opposed to more drought-sensitive traits in larger trees (Tables 1,5). As discussed above, there was little meaningful variation in traits with height at the community level. When $ln[H]$ and hydraulic traits were considered together in multivariate models, the effect of $ln[H]$ on $R$ was consistently negative (Table 5). Simultaneously in the 1999 univariate model, this co-consideration resulted in a non-signficant positive $ln[H]$-$R$ correlation (Table 4).

@@

Our results reject the hypothesis that overall community responses varied across droughts (*H3.1*). Within the context of mixed effects models, there were no significant differences in $R$ across drought years (Table 4). **see comment 36**

Our findings give marginal support for Hypothesis *H3.2*, whereby we predicted directions of responses varied across droughts. In the majority of cases, response directions were consistent across droughts in both univariate and multivariate models (Tables 1,4,5). However, there were a few exceptions-most commonly in the categorical variables ($CP$ and $XP$) but also for $PLA_{dry}$ in one multivariate model for the 1999 drought (Tables 4, 5). These differences may be spurious as opposed to statistically meaningful. Among the univariate models, there was no instance where predictor variables significantly improved the models of two different droughts (dAIC $\ge$ 2), but with contrasting coefficients (Table 4). Among the multivariate models, $CP$ was not consistently in the top models for any drought (Table 5), and $PLA_{dry}$ only appeared with a positive coefficient in the weakest predictor model of the six total models for the 1999 drought (contrasting with a negative coefficient in the univariate model; Table 4). It is probable that in the case of ring porosity, ring-porous species had lower $R$ than diffuse- and semi-ring- porous species in the 1977 drought, contrasting with higher resistance in 1966 and 1999 (Tables 4,5). Note, however, that $XP$ was not a signficant predictor on its own for the 1977 drought.

Further, our findings support the hypothesis (*H3.3*) that the strength of predictor variables was different across the droughts (Tables 1,4,5). For instance, $ln[H]$ and $PLA_{dry}$ had much stronger negative effects in 1966 than in the other two years, $ln[TWI]$ had the strongest negative effect in 1977, and $XP$ (higher $R$ among ring-porous trees) was strongest in 1999 (Tables 4,5). **see comment 37**

@@

Our results reveal how tree size, microenvironment, **see comment 38** and hydraulic traits shaped tree growth responses across three droughts in a temperate deciduous forest (Table 1). The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height [@liu_effect_1993; @stovall_tree_2019]. There was a marginal additional effect of crown exposure, with the highest drought impacts felt by the most exposed and the most suppressed trees. This is consistent with observations of both the former (e.g., [@suarez_factors_2004]; [@scharnweber_confessions_2019]) and the latter (**REFS**). **see comment 39** There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions experienced more growth reduction (consistent with [@zuleta_drought-induced_2017]), and larger rooting volume provided no advantage in the drier microenvironments. **see comment 40** The lower drought resistance of larger trees was not driven by a trend of more drought-sensitive species dominating the canopy. Drought sensitivity was also not consistently linked to species' $LMA$, $WD$, or xylem architecture. However, drought resistance was negatively correlated with the leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts [@scoffoni_leaf_2014;@bartlett_correlations_2016; @medeiros_extensive_2019]. This is a novel finding in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses. The direction of responses was mostly consistent across droughts, supporting the conclusion that they were driven by fundamental physiological mechanisms. However, the strengths of each predictor varied across droughts, indicating that specific drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected. These findings significantly advance our knowledge of the factors that confer vulnerability or resistance on trees during drought.

The droughts considered here were of similar severity (Fig. 1b) and fairly moderate; droughts of this magnitude have occurred with an average frequency of approximately one per 10-15 years (Fig. 1a, @helcoski_growing_2019). Therefore, we excpect that most species are adapted, and individual trees acclimatized, to survive droughts of this nature when not coupled with other disturbances. While the majority of trees experienced reduced growth, a substantial portion had increased growth (Fig. 1b), underlining the fact that these droughts did not induce extreme stress on the entire forest. It is likely for this reason, along with other non-climate factors influencing tree growth, that our best models characterize only a modest amount of variation: 11-13% for all droughts combined, and 21-26% for each individual drought (Table 5). Methodologically, the moderate nature of these droughts is an advantage because our analysis considers only trees that survived all of these droughts, and we lack information on the trees that were killed. These are likely to be relatively modest in number, and local forest monitoring data stretching back to the late 1980s confirms that the 1999 drought did not trigger major declines in tree abundance or biomass (*Anderson-Teixeira et al., in revision*). Thus, the droughts considered here are substantially weaker than those that have triggered massive tree die-off (e.g., [@allen_global_2010]), many of which have shaped our understanding about the role of tree size [@bennett_larger_2015; @stovall_tree_2019] and--to some extent--traits underlying drought response mechanisms (@greenwood_tree_2017; @anderegg_meta-analysis_2016). Nevertheless, our results are consistent with findings from more extreme droughts.
